---
title: "ADS 506 — Week 5 Shiny Wine Varietals"
author: "Katie Kimberling"
date: today
format:
  pdf:
    toc: false
    geometry: ["margin=1in"]
execute:
  echo: false
  warning: false
  message: false
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(fpp3)
library(gt)
library(readr)
```

## Tab 1 - Visualize Data
### Data imort / preparation
Inputs

* Multiselect for Wine Varietal
* Date range
* First date is train date cutoff - default to 1 year prio to the max date
* Second date is the end of the forecast horizon -- default ot last date

```{r}
# Read in data and convert to long format with better error handling
Aus_wine_raw <- read_csv(
  "AustralianWines.csv",
  col_types = cols(
    Month = col_character(),
    Fortified = col_double(),
    Red = col_double(),
    sparkling = col_double(),
    `Sweet white` = col_double(),
    `Dry white` = col_double(),
    Rose = col_double()
  ),
  na = c("", "NA", "*")  # Treat * as missing value
)

wine_data <- Aus_wine_raw |>
  mutate(Month = yearmonth(my(Month))) |>
  pivot_longer(
    cols = -Month,
    names_to = "Varietal",
    values_to = "Sales"
  ) |>
  as_tsibble(index = Month, key = Varietal)

# Define training cutoff - default to 1 year before max date
train_end <- max(wine_data$Month) - 12

# Split into training and validation
training <- wine_data |>
  filter(Month <= train_end)

validation <- wine_data |>
  filter(Month > train_end)
```

### Data visualization
Give the user the ability to select one or more purposes to visualize in our app
```{r}
# visualize the data with autoplot and facets by varietal
wine_data |>
    autoplot(Sales) +
    geom_vline(xintercept = as.Date(train_end), linetype = "dashed", color = "red", linewidth = 0.6) +
    labs(y = "Sales (kL)", title = "Australian Wine Sales by Varietal") +
    facet_wrap(vars(Varietal), ncol = 1, scales = "free_y") +
    theme_minimal()
```

### Optional: Seasonality / Decomposition view
```{r}
# STL decomposition for one varietal as example
wine_data |>
  filter(Varietal == "Red") |>
  model(stl = STL(Sales)) |>
  components() |>
  autoplot() +
  labs(title = "STL Decomposition: Red Wine Sales") +
  theme_minimal()
```

## Tab 2 - Model Building
Inputs:
* Toggle for model specification (default off)
* Toggle the training accuracy (default off)
* Forecast horizon input (default 12 months)


### Model fitting
Build multiple models for each Varietal using TSLM, ETS and ARIMA.
This is a second area or tab for our app
```{r}
# fit TSLM, ETS, and ARIMA models by varietal
fit <- training |>
    model(
        tslm = TSLM(Sales ~ trend() + season()),
        ets = ETS(Sales),
        arima = ARIMA(Sales)
    )
fit # output the model specification
```

### Training accuracy
```{r}
fit |>
    accuracy() |>
    select(Varietal, .model, RMSE, MAE, MAPE) |>
    arrange(.model, MAPE) |>
    gt() |>
    fmt_number(
        columns = c(RMSE, MAE, MAPE),
        decimals = 2
    ) |>
    tab_header(title = "Training Accuracy by Model and Varietal")
```

### Forcast accuracy
```{r}
# Generate forecasts for validation period (12 months)
fc <- fit |>
    forecast(h = 12)

fc |>
    accuracy(wine_data) |>
    select(Varietal, .model, RMSE, MAE, MAPE) |>
    arrange(.model, MAPE) |>
    gt() |>
    fmt_number(decimals = 2) |>
    tab_header(title = "Validation Accuracy by Model and Varietal")
```

## Tab 3
Iputs:
* User can select model for forecast visualization
* User can select which varietals to display


### Forecasts visualization
The third tab in my Shiny app - outputting visualizations of the forecasts
```{r}
# Show last 2 years of training data plus forecasts
trn_start <- yearmonth(as.Date(train_end) - years(2))

fc |>
    autoplot(wine_data |> filter(Month >= trn_start), level = c(80, 95)) +
    labs(y = "Sales (kL)", title = "Wine Sales Forecasts by Varietal and Model") +
    facet_grid(Varietal ~ .model, scales = "free_y") +
    geom_vline(xintercept = as.Date(train_end), linetype = "dashed", color = "red", linewidth = 0.6) +
    theme_minimal()
```
## Model Specifications Summary
* Exract and display model specification in a readable format
```{r}
# Extract ETS component forms
ets_specs <- fit |>
  filter(.model == "ets") |>
  mutate(ets_spec = format(ets)) |>
  select(Varietal, ets_spec)

# Extract ARIMA orders
arima_specs <- fit |>
  filter(.model == "arima") |>
  mutate(arima_spec = format(arima)) |>
  select(Varietal, arima_spec)

# Display as tables
ets_specs |>
  gt() |>
  tab_header(title = "ETS Model Specifications")

arima_specs |>
  gt() |>
  tab_header(title = "ARIMA Model Specifications")
